Disentangling linkages between satellite derived forest structure and productivity essential biodiversity variables.

### 2023-06-01 scripts/manuscript/manuscript.tex
secondary sampling scheme, revised manuscript for ncc

\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\usepackage{float}
\makeatletter
\let\oldlt\longtable
\let\endoldlt\endlongtable
\def\longtable{\@ifnextchar[\longtable@i \longtable@ii}
\def\longtable@i[#1]{\begin{figure}[H]
\onecolumn
\begin{minipage}{0.5\textwidth}
\oldlt[#1]
}
\def\longtable@ii{\begin{figure}[H]
\onecolumn
\begin{minipage}{0.5\textwidth}
\oldlt
}
\def\endlongtable{\endoldlt
\end{minipage}
\twocolumn
\end{figure}}
\makeatother
\journal{Remote Sensing of Environment}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Disentangling linkages between satellite derived forest structure and productivity across a range of forest types and ecosystems.},
  pdfauthor={Evan R. Muise; Nicholas C. Coops; Txomin Hermosilla; A. Cole Burton; Margaret E. Andrew; Stephen S. Ban},
  pdfkeywords={structural equation modelling, remote
sensing, landsat, forest structure, forest productivity},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Disentangling linkages between satellite derived forest structure
and productivity across a range of forest types and ecosystems.}
\author[1]{Evan R. Muise%
\corref{cor1}%
}
 \ead{evan.muise@student.ubc.ca} 
\author[1]{Nicholas C. Coops%
%
}
 \ead{nicholas.coops@ubc.ca} 
\author[2]{Txomin Hermosilla%
%
}
 \ead{txomin.hermosillagomez@NRCan-RNCan.gc.ca} 
\author[1]{A. Cole Burton%
%
}
 \ead{cole.burton@ubc.ca} 
\author[3]{Margaret E. Andrew%
%
}
 \ead{M.Andrew@murdoch.edu.au} 
\author[4]{Stephen S. Ban%
%
}
 \ead{Stephen.Ban@gov.bc.ca}

\affiliation[1]{organization={University of British Columbia, Forest
Resources Management},addressline={2424 Main Mall},city={Vancouver, BC,
Canada},postcode={V6T 1Z4},postcodesep={}}
\affiliation[2]{organization={Natural Resources Canada, Canada Forest
Service (Pacific Forestry Centre)},addressline={506 Burnside Rd
W},city={Victoria, BC, Canada},postcode={V8Z 1M5},postcodesep={}}
\affiliation[3]{organization={Murdoch University, Environmental and
Conservation Sciences and Harry Butler Institute},city={Murdoch, WA,
Australia},postcode={6150},postcodesep={}}
\affiliation[4]{organization={Ministry of Environment and Climate Change
Strategy, BC Parks},addressline={525 Superior Street},city={Victoria,
BC, Canada},postcode={V8V 1T7},postcodesep={}}

@@

\begin{abstract}
This is the abstract. Lorem ipsum dolor sit amet, consectetur adipiscing
elit. Vestibulum augue turpis, dictum non malesuada a, volutpat eget
velit. Nam placerat turpis purus, eu tristique ex tincidunt et. Mauris
sed augue eget turpis ultrices tincidunt. Sed et mi in leo porta
egestas. Aliquam non laoreet velit. Nunc quis ex vitae eros aliquet
auctor nec ac libero. Duis laoreet sapien eu mi luctus, in bibendum leo
molestie. Sed hendrerit diam diam, ac dapibus nisl volutpat vitae.
Aliquam bibendum varius libero, eu efficitur justo rutrum at. Sed at
tempus elit.
\end{abstract}

\begin{keyword}
    structural equation modelling \sep remote
sensing \sep landsat \sep forest structure \sep 
    forest productivity
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, sharp corners, interior hidden, frame hidden, enhanced, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

@@

Biodiversity is the summation of variation in biological life, across
genes, species, communities, and ecosystems. Currently, biodiversity is
in decline, facing extinction rates above the background extinction rate
\citep{thomas2004, urban2015}, and homogenization of communities at
various scales \citep{mcgill2015}. In response, the global biodiversity
community is making efforts to assess and halt the degradation of
biodiversity on earth. The Group for Earth Observation Biodiversity
Observation Network has developed the Essential Biodiversity Variables
\citep[EBVs,][]{pereira2013}, designed as an analog to the Essential
Climate Variables framework \citep{bojinski2014}. EBVs are designed to
be global in scope, relevant to biodiversity information, feasible to
utilize, and complementary to one another \citep{skidmore2021}.

There are six EBV classes, each corresponding to a different facet of
biodiversity, including, genetic composition, species populations,
species traits, community composition, ecosystem structure, and
ecosystem function \citep{pereira2013}. \citet{fernández2020} divides
the six classes into two approaches, with one focusing on species
biodiversity, and the other focusing on ecosystem diversity. Remote
sensing has proven to be capable of measure five of the six classes,
with the exception being genetic composition, which requires in-situ
observation and samples \citep{skidmore2021}. Notably, while remote
sensing can provide information on the remaining two species EBV
classes, this information is typically acquired with an \emph{ad-hoc}
approach. This collects high resolution, small spatial extent
information, rather than the global or regional scales required for
biodiversity assessment. Community composition falls into a similar
dilemma, requiring species population information which necessitates
high resolution spatial data.

The remaining two classes, ecosystem structure and function, are
incredibly well suited to be examined at global or regional scales using
mid-resolution satellite imagery, such as that provided by the Landsat
series of satellites. Advances in satellite remote sensing processing
have allowed 3d forest structure data to be imputed across wide spatial
scales \citep{matasci2018, coops2021} using data fusion approaches
involving collected lidar data and optical/radar data. Other advances in
image compositing have allowed yearly summaries of vegetation
productivity to be calculated at regional to global scales, summarizing
the yearly energy totals, minimums, and variations \citep{radeloff2019}.
These datasets correspond quite well with the EBV classes ecosystem
structure (forest structural diversity metrics), and ecosystem function
(forest productivity metrics).

Forest structural diversity has been linked to biodiversity at various
scales \citep{guo2017, bergen2009, gao2014}. Increased structural
complexity is hypothesized to create additional niches, leading to
increased species diversity \citep{bergen2009}. The relationship between
forest structure and biodiversity is commonly assessed using avian
species diversity metrics \citep{macarthur1961, goetz2007}, however,
other clades (and habitats), have been used
\citep{davies2014, nelson2005}. Many metrics derived from lidar remote
sensing have been used as local indicators of biodiversity, including
canopy cover, canopy height, vertical profiles, and aboveground biomass,
while other 2nd order derived metrics such as canopy texture, height
class distribution, edges, and patch metrics have been used to examine
habitat and biodiversity at landscape scales \citep{bergen2009}.

The dynamic habitat indices (DHIs) are indicators of productivity
calculated by summarizing vegetation indices over the course of one (or
multiple) years \citep{radeloff2019}. These indices have been related to
multiple facets of biodiversity at a range of scales, including species
occurrence and abundance \citep{razenkova2020}, alpha
\citep{radeloff2019} and beta diversity \citep{andrew2012}. Hypotheses
behind the biodiversity productivity relationships have been
established, including the species-energy hypothesis, the environmental
stress hypothesis, and the environmental stability hypothesis
\citep{coops2019}. The cumulative DHI calculates the total amount of
energy available in a given pixel over the course of a year. Cumulative
DHI is strongly linked to the available energy hypothesis, which
suggests that with greater available energy species richness will
increase \citep{wright1983}. The minimum DHI, which calculates the
lowest productivity over the course of a year can be matched to the
environmental stress hypothesis, which proposes that higher levels of
minimum available energy will lead to higher species richness
\citep{currie2004}. Finally, the variation DHI, which calculates the
coefficient of variance in a vegetation index through the course of a
year, corresponds to the environmental stability hypothesis which states
that lower energy variation throughout a year will lead to increased
species richness \citep{williams2008}.

Linkages between forest structure and productivity (namely, vegetation
indices) have been examined for nearly 20 years \citetext{\citealp[
\citet{knyazikhin1998}]{huete2002}; \citealp{myneni1994}}. While there
is significant theoretical and empirical evidence for their relationship
at single time points (within a single image) \citep{myneni1994},
various relationship directions and shapes have been found between
forest structure and productivity metrics \citep{ali2019}. These
relationships, their shapes, and their strengths have been attributed to
multiple possible hypotheses and can vary based on environmental
conditions \citep{ali2019}. The relationship between forest structural
diversity metrics and annual productivity summaries has yet to be fully
examined, including the DHIs.

Some vegetation structure metrics are simpler, and more accurate, to
calculate than others \citep{coops2021}. These basic metrics, such as
canopy height and canopy cover,can then be used to estimate additional
structure metrics, such as basal area or total biomass. Canopy height
and cover are commonly used as an indicator of vertical and horizontal
variation, respectively. Recently, more attention has been paid to
internal structural complexity metrics, which can be more difficult and
time consuming to generate \citep{coops2021, ma2022} .

In this study, we seek to untangle the relationship between two EBVs:
forest structure diversity metrics and yearly summaries of forest
productivity. To accomplish this, we assess this relationship using path
analysis to assess the direct and indirect (as mediated by more complex
forest structural diversity metrics) effects of commonly collected
forest structural metrics (canopy height and canopy cover) on yearly
productivity summaries. Doing so will in turn assess their utility as
complementary EBVs. We ran this analysis separately across four forested
land covers and the forested ecosystems of British Columbia, Canada.

@@

\hypertarget{study-area}{%
\subsection{Study Area}\label{study-area}}

British Columbia is the westernmost province of Canada, and is home to a
variety of terrestrial ecosystems \citep{pojar1987}. Approximately 64\%
of the province is forested \citep{bcministryofforests2003}. There is a
large amount of ecosystem variation in the province, with large climate
and topographic gradients. The Biogeoclimatic Ecosystem Classification
(BEC) system identifies 16 zones based on the dominant tree species and
the ecosystems general climate. These zones can be further split into
subzones, variants, and phases based on microclimate, precipitation, and
topography.

@@

@@

@@

@@

>>>>

\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\usepackage{float}
\makeatletter
\let\oldlt\longtable
\let\endoldlt\endlongtable
\def\longtable{\@ifnextchar[\longtable@i \longtable@ii}
\def\longtable@i[#1]{\begin{figure}[H]
\onecolumn
\begin{minipage}{0.5\textwidth}
\oldlt[#1]
}
\def\longtable@ii{\begin{figure}[H]
\onecolumn
\begin{minipage}{0.5\textwidth}
\oldlt
}
\def\endlongtable{\endoldlt
\end{minipage}
\twocolumn
\end{figure}}
\makeatother
\journal{Remote Sensing of Environment}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Disentangling linkages between satellite derived forest structure and productivity essential biodiversity variables.},
  pdfauthor={Evan R. Muise; Margaret E. Andrew; Nicholas C. Coops; Txomin Hermosilla; A. Cole Burton; Stephen S. Ban},
  pdfkeywords={remote sensing, landsat, forest structure, forest
productivity, dynamic habitat indices, essential biodiversity
variables},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Disentangling linkages between satellite derived forest structure
and productivity essential biodiversity variables.}
\author[1]{Evan R. Muise%
\corref{cor1}%
}
 \ead{evan.muise@student.ubc.ca} 
\author[2]{Margaret E. Andrew%
%
}
 \ead{M.Andrew@murdoch.edu.au} 
\author[1]{Nicholas C. Coops%
%
}
 \ead{nicholas.coops@ubc.ca} 
\author[3]{Txomin Hermosilla%
%
}
 \ead{txomin.hermosillagomez@NRCan-RNCan.gc.ca} 
\author[1]{A. Cole Burton%
%
}
 \ead{cole.burton@ubc.ca} 
\author[4]{Stephen S. Ban%
%
}
 \ead{Stephen.Ban@gov.bc.ca}

\affiliation[1]{organization={University of British Columbia, Forest
Resources Management},addressline={2424 Main Mall},city={Vancouver, BC,
Canada},postcode={V6T 1Z4},postcodesep={}}
\affiliation[2]{organization={Murdoch University, Environmental and
Conservation Sciences and Harry Butler Institute},city={Murdoch, WA,
Australia},postcode={6150},postcodesep={}}
\affiliation[3]{organization={Natural Resources Canada, Canada Forest
Service (Pacific Forestry Centre)},addressline={506 Burnside Rd
W},city={Victoria, BC, Canada},postcode={V8Z 1M5},postcodesep={}}
\affiliation[4]{organization={Ministry of Environment and Climate Change
Strategy, BC Parks},addressline={525 Superior Street},city={Victoria,
BC, Canada},postcode={V8V 1T7},postcodesep={}}

@@

\begin{abstract}
{[}insert abstract here - nicholas ignore the formatting on the authors
for now, i can output to pdf and have it formatted nicely for
submission{]}
\end{abstract}

\begin{keyword}
    remote sensing \sep landsat \sep forest structure \sep forest
productivity \sep dynamic habitat indices \sep 
    essential biodiversity variables
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, interior hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, frame hidden, sharp corners]}{\end{tcolorbox}}\fi

@@

With biodiversity being in decline, and facing extinction rates above
the background extinction rate \citep{thomas2004, urban2015}, as well as
the homogenization of communities at various scales \citep{mcgill2015}
it is integral to be able to monitor how biodiversity is changing across
the globe. In response, the global biodiversity community is making
efforts to assess and halt the degradation of biodiversity. The Group
for Earth Observation Biodiversity Observation Network has developed the
Essential Biodiversity Variables \citep[EBVs,][]{pereira2013}, designed
as an analog to the Essential Climate Variables framework
\citep{bojinski2014}. EBVs are designed to be global in scope, relevant
to biodiversity information, feasible to use, and complementary to one
another \citep{skidmore2021}. While it can be incredibly difficult, time
consuming, and expensive to collect data on biodiversity across wide
swaths of land and varying ecosystems, EBVs, which can be correlated to
sampled biodiversity information, allow for the monitoring and
asssessment of protected area effectiveness and ecosystem health at
large spatial scales \citep{hansen2021}. There are six EBV classes, each
of which correspond to a different facet of biodiversity including
species populations, species traits, community composition, ecosystem
structure, ecosystem function, and genetic composition
\citep{pereira2013}.

Satellite remote sensing has proven to be capable of measuring five of
the six EBV classes, the exception being genetic composition, which
requires in-situ observation and sampling \citep{skidmore2021}. Species
populations - and in turn community composition - can be assessed with
very-high-resolution imagery to identify tree species at the tree-crown
scale, however it is difficult and computationally expensive to extend
these analyses to broader extents \citep{fassnacht2016, graves2016}
while species traits such as vegetation phenology have been observed at
the single-tree scale using, for example, PlanetScope imagery and
drone-based measurements \citep{wu2021}. However, the spatially limited
and often \emph{ad-hoc} data collection approaches associated with
monitoring individuals is not conducive to the global or regional scales
required for biodiversity trend assessment \citep{valdez2023}.

The two landscape-level EBVs (ecosystem structure and function) are well
suited to be examined at large spatial scales using coarser spatial
measurements, such as those taken from satellites by the Moderate
Resolution Imaging Spectroradiometer {[}MODIS; \citet{zhang2003}{]}, the
Landsat imaging systems \citep{fisher2006}, or Sentinel-2
\citep{helfenstein2022, darvishzadeh2019} programs. These mid-resolution
satellites can monitor processes at broader extents but the coarse
spatial resolution removes the ability to relate these traits to
individual organisms. As a result, satellite remote sensing data has
been shown to be arguably the most effective at monitoring ecosystem
based EBVs focused on structure and function. These EBVs classes can be
monitored at regional to global extents through the use of optical
imagery \citep{cohen2004}, as well as active sensors such as lidar
(light detection and ranging) and radar
\citep{guo2017, lefsky2002, lang2021, neuenschwander2019, coops2016}.

\hypertarget{ebv---ecosystem-structure}{%
\subsection{EBV - Ecosystem Structure}\label{ebv---ecosystem-structure}}

Forest structural diversity has been linked to biodiversity at various
scales \citep{guo2017, bergen2009, gao2014}. Structural attributes range
in complexity from simple (canopy cover; canopy height), to more complex
(vertical and horizontal structural complexity) to modelled (aboveground
biomass; basal area), all of which can be assessed using lidar data
\citep{coops2021}. A suite of these lidar-derived attributes have been
used as local indicators of biodiversity, including simple metrics such
as canopy cover and canopy height as well as derived metrics including
vertical profiles, aboveground biomass
\citep{lefsky1999, guo2017, coops2016}. Other second order derived
metrics such as canopy texture, height class distribution, edges, and
patch metrics have also been used to examine habitat and biodiversity at
landscape scales \citep{bergen2009}. Advances in satellite remote
sensing processing have allowed 3D forest structure data to be imputed
across wide spatial scales \citep{matasci2018, coops2021} using data
fusion approaches involving collected lidar data and optical/radar data.

Increased forest structural complexity has been hypothesized to create
additional niches, leading to increased species diversity
\citep{bergen2009}, which has been frequently demonstrated using avian
species diversity metrics \citep{macarthur1961}. For example:
\citet{herniman2020} used spectral and lidar derived forest structure
data to model avian habitat suitability; \citet{clawges2008} found that
lidar derived forest structural attributes are capable of identifying
habitat types associated with avian species in pine/aspen forests;
\citet{goetz2007} used canopy structural diversity to predict bird
species richness, finding that canopy vertical distribution was the
strongest predictor of species richness. Forest structural metrics have
also been used to study biodiversity in other clades as well
\citep{davies2014, nelson2005}.

\hypertarget{ebv---ecosystem-function}{%
\subsection{EBV - Ecosystem Function}\label{ebv---ecosystem-function}}

With respect to ecosystem function, energy availability in an ecosystem
has shown to be a predictor of species richness and abundances at
various scales \citep{chase2002, radeloff2019, coops2019, razenkova},
and is measurable using satellite remote sensing via the use of various
vegetation indices \citep{huete2002, radeloff2019}. Vegetation indices,
which are indicative of photosynthetic activity, are commonly used as
proxies of gross primary productivity \citep{huang2019}. These
vegetation indices have also been used to assess patterns in
biodiversity at single time points \citep{bonn2004}, and more recently,
through yearly summaries of productivity
\citep{berry2007, radeloff2019}. The relationship between energy
availability and biodiversity occurs via various hypothesized
mechanisms, such as the available energy hypothesis
\citep{currie2004, wright1983}, the environmental stress hypothesis
\citep{currie2004}, and the environmental stability hypothesis
\citep{williams2008}. These three hypotheses have in turn been linked to
patterns of annual surface reflectance in remote sensing data
\citep{berry2007, radeloff2019}.

\citet{berry2007} first explored this idea by proposing the linkage of
intra-annual summaries of MODIS derived GPP to dispersive bird species.
This idea was further refined into the Dynamic Habitat Indices {[}DHIs;
\citet{coops2008}{]}, which have now been shown to be well suited to
assess the three aforementioned hypotheses at global scales
\citep{radeloff2019}. The cumulative DHI calculates the total amount of
energy available in a given pixel over the course of a year. Cumulative
DHI is strongly linked to the available energy hypothesis, which
suggests that with greater available energy species richness will
increase \citep{wright1983}. The minimum DHI, which calculates the
lowest productivity over the course of a year can be matched to the
environmental stress hypothesis, which proposes that higher levels of
minimum available energy will lead to higher species richness
\citep{currie2004}. Finally, the variation DHI, which calculates the
coefficient of variance in a vegetation index through the course of a
year, corresponds to the environmental stability hypothesis which states
that lower energy variation throughout a year will lead to increased
species richness \citep{williams2008}.

\hypertarget{biodiversity-monitoring-with-ebvs}{%
\subsection{Biodiversity Monitoring with
EBVs}\label{biodiversity-monitoring-with-ebvs}}

Biodiversity monitoring programs often require a range of information in
order to accurately assess changes in ecosystem integrity
\citep{lindenmayer2010}. Choosing datasets that are most closely related
to the phenomenon of interest in a given application allows for direct
connections to monitoring results and management actions
\citep{pressey2021}. With the advent of large-extent monitoring methods
like satellite remote sensing, and a proliferation of potential EBVs
datasets, it becomes important to assess the interrelationships between
these datasets, and assess their complementary of the information to
reduce duplication of efforts \citep{pereira2013, skidmore2021}. When
strong relationships are present between EBV classes, it becomes
possible to assess the ecological relationships between potential EBVs.
On the other hand, when datasets do not appear related, they may be well
suited to be used in monitoring programs together, as complementary
EBVs.

Linkages between forest ecosystem structure and function have been
examined within a remote sensing context for over 20 years
\citetext{\citealp[
\citet{knyazikhin1998}]{huete2002}; \citealp{myneni1994}}. While there
is significant theoretical and empirical evidence for their relationship
at single time points (within a single image) \citep{myneni1994},
various relationship directions and shapes have been found between
forest structure and function metrics \citep{ali2019}. Hypothesized
mechanisms such as niche complementary have shown that aboveground
biomass increases with stand structure \citep{zhang2012}, while
asymmetric competition for light can reduce forest productivity with
increased structural complexity \citep{bourdier2016}. The relationship
in particular between forest structural diversity metrics - which are
now more accurately and comprehensively derived from lidar data - and
temporal variation in functional metrics, specifically the metrics of
ecosystem productivity via the DHI framework, have yet to be fully
examined.

The overall goal of this paper is to assess patterns of forest ecosystem
structure and function and their complementarity across a wide range of
ecosystems encompassing significant environmental gradients. To do so,
we synthesize data from moderate-scale remote-sensing derived metrics of
forest structure, represented as both simple ALS-extracted metrics of
canopy height, cover and vertical complexity, as well as modelled forest
structure attributes including volume, and aboveground biomass, with a
well-established remote sensing derived index on ecosystem function. Our
first question is to examine how ecosystem structure and function
complement one another across a large environmental gradient and then
compare the simple and modelled representations of forest structure to
different levels of ecosystem function. This question is important as it
provides insights to the EBV community around complementarity of remote
sensing metrics when describing the structure and function of ecosystems
and proposes a method to examine potential overlap when generating
remote sensing EBVs.

Our second question examines the independent and shared relationships of
ecosystem structure height and cover, with modelled forest structure, on
ecosystem function. This provides insight into the choice of remote
sensing attributes to use when developing EBVs within a single EBV
class. Remote sensing datasets can comprise relatively unprocessed
observations, in this case ALS measures of height and cover which are
derived from the raw 3D point cloud vs modelled attributes, such as
biomass and volume, which involve the use the statistical relationships
with field data to transform the observations into more refined data
products. Assessing which of these two (or combination of the two)
approaches has stronger or weaker correlations with estimates of
function provides insights into the choice of data used to build EBVs.
Lastly, we examine how the primary and modelled structure attributes
partition the variance of the DHIs within key biomes and forest types
across a large environmental range, examining to what extent ecosystem
and forest types impacts these relationships and thus providing insight
into the applicability of these results globally.

@@

\hypertarget{sec-study-area}{%
\subsection{Study Area}\label{sec-study-area}}

British Columbia is the westernmost province of Canada, and is home to a
variety of terrestrial ecosystems. Approximately 64\% of the province is
forested, with large environmental and topographic gradients
\citep{pojar1987, bcministryofforests2003}. The Biogeoclimatic Ecosystem
Classification (BEC) system identifies 16 zones based on the dominant
tree species and the ecosystems general climate. These zones can be
further split into subzones, variants, and phases based on microclimate,
precipitation, and topography \citep{pojar1987}. To examine trends
across the large environmental gradients, we group the BEC zones into
five broad biomes, specifically, the south interior, northern, montane,
alpine, and coastal groups similar to \citet{hamann2006}.

\{insert figure showing ecosystems of interest and grouped bec zones\}
need to remake figure showing ecosystems

\{table showing elevations and climate ranges for each included bec
zone\}

@@

\hypertarget{forest-structure}{%
\subsubsection{Forest Structure}\label{forest-structure}}

We used a suite of forest structural attributes (canopy height, canopy
cover, Lorey's height, overstory cover, basal area, aboveground biomass,
gross stem volume, mean elevation, elevation standard deviation, and
structural complexity {[}coefficient of variation in elevation
returns{]}). This dataset was created at a 30 m spatial resolution
according to \citet{matasci2018}. In brief, the method used a set of
lidar collections and field plots across Canada, and imputed the
remaining pixels using a random forest k-Nearest Neighbour approach on
Landsat-derived surface reflectance and auxiliary data such as
topography. Detailed information on the creation of this dataset can be
found in \citet{matasci2018}.

\hypertarget{dynamic-habitat-indices}{%
\subsubsection{Dynamic Habitat Indices}\label{dynamic-habitat-indices}}

We use an established set of indices of annual productivity shown to be
related to global biodvierstiy trends: the Dynamic Habitat Indices
{[}\citet{radeloff2019}{]}. The DHIs are a set of satellite remote
sensing derived productivity variables that summarize the cumulative
amount of available energy, the minimum available energy, and the
variation in available energy throughout a given year
{[}\citet{berry2007}; \citet{radeloff2019}{]}. The DHIs have previously
been produced at a global extent using MODIS imagery , and have been
used to assess alpha \citep{radeloff2019} and
beta{[}\citet{andrew2012}{]} diversity, species abundances
{[}\citet{razenkova}{]}, and construct novel ecoregionalizations
\citep{coops2009, andrew2013} . Recent studies have began to examine how
these indices can be constructed at a finer spatial resolution by using
multi-annual Landsat imagery to generate a single synthetic year of
monthly observations {[}\citet{razenkova2022}{]}.

The DHIs were calculated according to {[}\citet{razenkova2022};
RAZENKOVA LANDSAT DHI PAPER IN PRESS{]} for all of terrestrial British
Columbia. In brief, Google Earth Engine \citep{gorelick2017} was used to
obtain all valid Landsat pixels for a given study area, filtering out
pixels containing shadows, slouds, and cloud shadows within each image
\citep{zhu2012}, then calculated the NDVI for each pixel in each image.
They then calculated the median NDVI value for each month across the ten
year time span (2011-2020) to generate a synthetic year of monthly data.
The sum, minimum, and coefficient of variation across this synthetic
year of NDVI values is then calculated. More detailed information can be
found in \{Razenkova In Press\}.

@@

We conducted model-based sampling across the sixteen forest-dominated
ecosystems found within British Columbia {[}BC figure reference, I have
not made this figure yet{]}. Samples were randomly selected within each
BEC zone, in undisturbed pixels. Each sampled pixel had to have a
forested land cover class (coniferous, deciduous, mixed-wood, or
wetland-treed), and be surrounded by the same land cover class. The land
cover mask was generated by \citet[need to add to
zotero]{hermosilla2022} using a best-available-pixel composite, and an
inverse-distance weighted random forest approach across Canada.
Additionally, each pixel had to have a coefficient of variation less
than 0.5 in surrounding pixels in the two simplest forest structural
attributes, canopy height and canopy cover. A maximum of 3000 samples
were sampled in each BEC zone with a 1 km minimum sampling distance to
reduce the effects of spatial autocorrelation. All variables were
natural-log transformed and standardized. Variables containing zeros
were natural-log plus one transformed. Sampling was conducted in R
\citep{R-base} version 4.2.2 using the \textbf{sgsR} package
\citep{R-sgsR}. Focal analyses for the land cover classes and
coefficient of variation were calculated in Python version 3.9.

@@

\hypertarget{redundancy-analysis-and-variation-partitioning}{%
\subsubsection{Redundancy Analysis and Variation
Partitioning}\label{redundancy-analysis-and-variation-partitioning}}

Redundancy analysis (RDA) and variation partitioning were used to relate
the primary and modelled forest structure attributes to ecosystem
function across a broad environment range. Redundancy analysis functions
similarly to a multiple linear regression, except it is capable of
predicting multiple response variables. It accomplishes this by first
running a multiple linear regression of each predictor variable on each
response variable, then running a principle component analysis on the
residuals from each multiple linear regression. This reduces the
dimensionality of the output, and allows the relationship strength to be
assessed by calculating the loadings of both predictor and response
variables on the RDA axes. Partial redundancy analysis function
similarly, except also considers co-variates \citep{legendre2012}.
Redundancy analysis has widely been used in community ecology where
environmental variables of interest are compared to species composition
\citep{blanchet2014, kleyer2012}. While RDA is common in the ecological
literature, this analysis represents one of the first times this
technique has been applied to assess the complimentary of proposed EBVs.
Following the RDA, we employ ANOVAs to determine which axes are
significant, and calculate the proportion of variance attributable to
each axis using the eigenvalues generated from the RDA. We calculate
axis loadings for both predictor and response variables by calculating
the correlation between the variables and the RDA axes. Axis loadings
represent the relationship between a given variable and the RDA axis. We
only consider and display significant axes. To visualize the RDA for
both predictor and response variables, we display the results as path
diagrams, with loadings from each predictor to the RDA axis to the
response variables. The variance explained by each axis is also
displayed in the RDA box Figure~\ref{fig-rda-var}. All RDA calculations
were done in R \citep{R-base} version 4.2.2 using the \textbf{vegan}
package \citep{R-vegan}.

Variation partitioning is an extension of partial RDA which can assess
the overlap between the explanatory power of two datasets by utilizing
multiple partial RDAs and exchanging which datasets are considered the
predictor, and which is considered the co-variate \citep{legendre2012}.
Variation partitioning is traditionally displayed using a Venn diagram,
in which the percentage of variance explained by each dataset is in a
circle, and the overlap between circles represents the overlap in
variance explained. All variation partition analyses were done in R
\citep{R-base} version 4.2.2 using the \textbf{vegan} package
\citep{R-vegan}.

RDA and variation partitioning analyses were conducted for all samples,
as well as individually across each BEC zone and forest type. The
results were aggregated to BEC zone groups see~\ref{sec-study-area} for
visualization.

All code associated with the processing and analysis is available at
https://github.com/emuise/code-structProdSem.

@@

\begin{figure}

{\centering \includegraphics[width=2.66in,height=\textheight]{../../outputs/all_data_figure.drawio.png}

}

\caption{\label{fig-rda-var}A) Axis loadings from redundancy analysis of
primary and modelled forest structural attributes on the dynamic habitat
indices. B) Results from variation partitioning of primary and modelled
forest structural attributes on the DHIs. Both visualized analyses are
across all collected samples. See supplementary information for results
from each BEC zone and forest type.}

\end{figure}

Figure~\ref{fig-rda-var} A shows the results from redundancy analysis of
forest structural attributes on the dynamic habitat indices across the
entire sampled dataset. While there are three RDA axes associated with
the full dataset, the third axis explains 0.05\% of the variance in the
DHIs, and as such we do not show it. The first axis is strongly
represents all the DHIs (loadings \textgreater{} 0.85 for all DHIs), and
has the strongest loadings from canopy cover, basal area, aboveground
biomass, and gross stem volume. The other input attributes (canopy
height, structural complexity, and Lorey's height) have smaller
loadings. The second axis primarily represents the seasonality (Minimum
and Variation DHI) of the DHIs, with a very small loading on the
Cumulative DHI, and is primarily driven by canopy cover and complexity
(Figure~\ref{fig-rda-var} A).
